To log or not to log

### 2022-08-02 analysis/Analysis ECDC forecasts.Rmd
update manuscript

@@

Figure that shows Variance vs. mean of the observations. Makes the point that the
data looks neg. binomial. 
```{r}
# plot with mean vs. variance --------------------------------------------------
hub_data |>
  dplyr::select(location, target_type, target_end_date, true_value) |>
  unique() |>
  group_by(location, target_type) |>
  summarise(mean_value = mean(true_value), 
            sd_value = sd(true_value), 
            var_value = var(true_value)) |>
  ggplot(aes(y = var_value, x = mean_value)) +
  geom_point() + 
  facet_wrap(~ target_type, scales = "free") + 
  scale_x_continuous(trans = "log10", labels = label_fn) +
  scale_y_continuous(trans = "log10", labels = label_fn) +
  # scale_x_continuous(labels = label_fn) +
  # scale_y_continuous(labels = label_fn) +
  labs(y = "Variance observed value", x = "Mean observed value") + 
  geom_function(fun = function(x) {x + x^2}, linetype = "dashed", size = 0.5) + 
  theme_scoringutils() + 
  geom_smooth(alpha = 0.2, size = 0.2)

ggsave("output/figures/HUB-mean-var-obs.png", width = 7, height = 2.5)

@@

@@

diffs <- summarised_pairwise |>
  mutate(model = factor(model, levels = ranking_natural)) |>
  select(model, relative_skill, scale) |>
  pivot_wider(names_from = scale, values_from = relative_skill) |>
  mutate(model = factor(model, levels = ranking_natural), 
         difference = natural - log)

@@

```{r}
# correlation between values
hub_data |>
  select(location, target_end_date, target_type, true_value) |>
  unique() |>
  mutate(log_obs = log(pmax(true_value, 0) + 1)) |>
  group_by(target_type) |>
  summarise(cor = cor(true_value, log_obs))

# correlation between scores
cor_scores <- scores |>
  filter(horizon <=4) |>
  select(interval_score, scale, forecast_date, model, horizon, location, target_type) |>
  pivot_wider(names_from = scale, values_from = interval_score) |>
  group_by(horizon, target_type) |>
  summarise(cor = cor(natural, log), 
            spearman = cor(natural, log))

cor_scores |> 
  mutate(`Target type` = target_type) |>
  ggplot(aes(x = horizon, y = cor, 
             fill = `Target type`, color = `Target type`)) + 
  geom_histogram(stat = "identity", position = "dodge") +
  theme_scoringutils() + 
  coord_cartesian(ylim = c(0, 1)) +
  labs(y = "Correlation (scores natural vs. log)", x = "Forecast horizon")

# Correlation between ranks when going from natural to log
correlation_ranks <- rankings_wide |>
  filter(horizon < 5) |>
  group_by(horizon, target_type) |>
  summarise(cor = cor(natural, log)) |>
  rename(`Target type` = target_type)

plot_cor_ranks <- correlation_ranks |>
  ggplot(aes(x = horizon, y = cor, 
             fill = `Target type`, color = `Target type`)) + 
  geom_histogram(stat = "identity", position = "dodge") +
  theme_scoringutils() + 
  coord_cartesian(ylim = c(0, 1)) +
  labs(y = "Correlation (rank natural vs. log)", x = "Forecast horizon")

ggsave(filename = "output/figures/HUB-cor-ranks.png", width = 7, height = 4.5, 
       plot = plot_cor_ranks)

# Correlation between pairwise comparisons
correlation_rel_skill <- scores |>
  filter(horizon < 5) |>
  summarise_scores(
    by = c("model", "scale", "horizon", "target_type"), 
    relative_skill = TRUE) |>
  select(model, horizon, target_type, relative_skill, scale) |>
  pivot_wider(values_from = relative_skill, names_from = scale) |>
  group_by(horizon, target_type) |>
  summarise(cor = cor(natural, log))

plot_cor_skill <- correlation_rel_skill |>
  rename(`Target type` = target_type) |>
  ggplot(aes(x = horizon, y = cor, fill = `Target type`, color = `Target type` )) + 
  # geom_line() +
  # geom_point() +
  geom_histogram(stat = "identity", position = "dodge") +
  theme_scoringutils() + 
  coord_cartesian(ylim = c(0, 1)) +
  labs(y = "Correlation (rel. skill natural vs. log)", x = "Forecast horizon")

ggsave(filename = "output/figures/HUB-cor-rel-skill.png", width = 7, height = 4.5, 
       plot = plot_cor_skill)

plot_cors <- plot_cor_ranks + plot_cor_skill +
  plot_layout(guides = "collect") & 
  theme(legend.position = "bottom")

ggsave(filename = "output/figures/HUB-cors-skill-rank.png", width = 7, height = 3, 
       plot = plot_cors)

@@

```{r}
wis_vs_total <- function(scores, summary_fct = mean, fulldata = hub_data) {

  plot_df <- scores |> 
    filter(model == "EuroCOVIDhub-ensemble", 
           horizon == 2) |>
    group_by(target_type, location, scale) |>
    summarise(wis = summary_fct(interval_score), 
              .groups = "drop_last") |> 
    inner_join(total_values)

  plot_df |>
    ggplot(aes(y = wis, x = total)) + 
    geom_point() + 
    facet_wrap(~ target_type + scale, scale = "free") + 
    theme_scoringutils() + 
    theme(legend.position = "none") + 
    scale_x_continuous(labels = scale_fn) +
    scale_y_continuous(labels = scale_fn) +
    labs(y = "REPLACE", x = "Total number of observed values")
}

wis_vs_total(scores) + 
  labs(y = "Mean WIS")
ggsave("output/figures/HUB-mean-scores-vs-total.png", width = 7, height = 5)

wis_vs_total(scores, summary_fct = stats::sd) + 
  labs(y = "Sd WIS")
ggsave("output/figures/HUB-sd-scores-vs-total.png", width = 7, height = 5)

# log-log versions of that plot
wis_vs_total(scores) + 
  labs(y = "Mean WIS") + 
  scale_x_continuous(labels = scale_fn, trans = "log10") +
  scale_y_continuous(labels = scale_fn, trans = "log10") 
ggsave("output/figures/HUB-mean-scores-vs-total-log-log.png", width = 7, height = 5)

wis_vs_total(scores, summary_fct = stats::sd) + 
  labs(y = "Sd WIS") + 
  scale_x_continuous(labels = scale_fn, trans = "log10") +
  scale_y_continuous(labels = scale_fn, trans = "log10") 
ggsave("output/figures/HUB-sd-scores-vs-total-log-log.png", width = 7, height = 5)

# correlation

>>>>

```{r}

check <- hub_data |> 
  check_forecasts()

filtermodels <- check$unique_values |>
  filter(horizon >= 4, 
         target_type == 2,
         true_value > 900,
         quantile == 23, 
         forecast_date > 16,
         location == 32) |>
  pull(model)

scores <- scores[model %in% filtermodels]
scores[, type_and_scale := paste0(target_type, " - ", scale)]
scores[, type_and_scale := factor(type_and_scale, 
                                  levels = c("Cases - natural", "Deaths - natural",
                                             "Cases - log", "Deaths - log"))]

hub_data <- hub_data[model %in% filtermodels]
hub_data[, true_value := pmax(true_value, 0)]

@@

Figure 5? 
that shows Variance vs. mean of the observations. Makes the point that the
data looks neg. binomial. 
```{r}
# plot with mean vs. variance --------------------------------------------------
plot_var_mean <- function(hub_data, log = FALSE) {
  df <- hub_data |>
    dplyr::select(location, target_type, target_end_date, true_value) |>
    unique() |>
    group_by(location, target_type)

  f <- function(x) {x + x^2}

  if (log) {
    df <- mutate(df, true_value = log(pmax(true_value, 0) + 1))

    f <- function(x) {log(x + x^2 + 1)}
  }

  df |>
  summarise(mean_value = mean(true_value), 
            sd_value = sd(true_value), 
            var_value = var(true_value)) |>
  ggplot(aes(y = var_value, x = mean_value)) +
  geom_point() + 
  facet_wrap(~ target_type, scales = "free") + 
  scale_x_continuous(trans = "log10", labels = label_fn) +
  scale_y_continuous(trans = "log10", labels = label_fn) +
  labs(y = "Variance observed value", x = "Mean observed value") + 
  geom_function(fun = f, linetype = "dashed", size = 0.5) + 
  theme_scoringutils() + 
  geom_smooth(alpha = 0.2, size = 0.2)
}

p1 <- plot_var_mean(hub_data)
p2 <- plot_var_mean(hub_data, log = TRUE)

wis_vs_total <- function(scores, summary_fct = mean, fulldata = hub_data) {

  plot_df <- scores |> 
    filter(model == "EuroCOVIDhub-ensemble", 
           horizon == 2) |>
    group_by(location, target_type, scale, type_and_scale) |>
    summarise(wis = summary_fct(interval_score), 
              .groups = "drop_last") |> 
    inner_join(total_values)

  plot_df |>
    ggplot(aes(y = wis, x = total)) + 
    geom_point() + 
    facet_wrap(~ type_and_scale, scale = "free") + 
    theme_scoringutils() + 
    theme(legend.position = "none") + 
    scale_x_continuous(labels = scale_fn) +
    scale_y_continuous(labels = scale_fn) +
    labs(y = "REPLACE", x = "Total number of observed values")
}

# log-log versions of that plot

p3 <- wis_vs_total(scores) + 
  labs(y = "Mean WIS") + 
  scale_x_continuous(labels = scale_fn, trans = "log10") +
  scale_y_continuous(labels = scale_fn, trans = "log10")

p1 / p2 / p3 + 
  plot_layout(heights = c(1, 1, 2))

ggsave("output/figures/HUB-mean-scores-vs-total-log-log.png", width = 7, height = 6)

# ggsave("output/figures/HUB-mean-scores-vs-total-log-log.png", width = 7, height = 4)

# ggsave("output/figures/HUB-mean-var-obs.png", width = 7, height = 2.5)
```

@@

```{r}
v1 <- c("a","b","c","d","e","f","g")
v2 <- c("b","x","e","c","z","d","a")

o <- 0.05
DF <- data.frame(x = c(rep(1, length(v1)), rep(2, length(v2))),
                 x1 = c(rep(1 + o, length(v1)), rep(2 - o, length(v2))),
                 y = c(rev(seq_along(v1)), rev(seq_along(v2))),
                 g = c(v1, v2))

library(ggplot2)
library(grid)
ggplot(DF, aes(x=x, y=y, group=g, label=g)) +
  geom_path(aes(x=x1), arrow = arrow(length = unit(0.02,"npc")), 
            size=1, color="green") +
  geom_text(size=10) +
  theme_minimal() +
  theme(axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank())
```

@@

diffs <- summarised_pairwise |>
  mutate(model = factor(model, levels = ranking_natural)) |>
  select(model, relative_skill, scale) |>
  pivot_wider(names_from = scale, values_from = relative_skill) |>
  mutate(model = factor(model, levels = ranking_natural), 
         difference = log - natural)

@@

```{r}

# 1. observations natural vs. log --> pearson
# 2. correlation scores natural vs. log (single forecasts) --> spearman

# Table
# correlation between values
cor_data <- hub_data |>
  select(location, target_end_date, target_type, true_value) |>
  unique() |>
  mutate(log_obs = log(pmax(true_value, 0) + 1)) |>
  group_by(target_type) |>
  summarise(cor = cor(true_value, log_obs))

# # correlation between scores
# # not really meaningful: if we do ranks it's just going to be ranks over everything which doesn't make sense
# # if we do individual scores than it's a bit unclear what the "correct" correlation should be
# cor_scores <- scores |>
#   filter(horizon <=4) |>
#   select(interval_score, scale, forecast_date, model, horizon, location, target_type) |>
#   pivot_wider(names_from = scale, values_from = interval_score) |>
#   group_by(horizon, target_type) |>
#   summarise(cor = cor(natural, log), 
#             spearman = cor(natural, log, method = "spearman"))

# Correlation between ranks when going from natural to log
correlation_ranks <- rankings_wide |>
  filter(horizon < 5) |>
  group_by(horizon, target_type) |>
  summarise(cor = cor(natural, log), 
            spearman = cor(natural, log, method = "spearman")) |>
  rename(`Target type` = target_type)

# Correlation between pairwise comparisons
correlation_rel_skill <- scores |>
  filter(horizon < 5) |>
  summarise_scores(
    by = c("model", "scale", "horizon", "target_type"), 
    relative_skill = TRUE) |>
  select(model, horizon, target_type, relative_skill, scale) |>
  pivot_wider(values_from = relative_skill, names_from = scale) |>
  group_by(horizon, target_type) |>
  summarise(cor = cor(natural, log), 
            spearman = cor(natural, log, method = "spearman"))

p_cor_ranks <- correlation_ranks |>
  ggplot(aes(x = horizon, y = cor, 
             fill = `Target type`, color = `Target type`)) + 
  geom_histogram(stat = "identity", position = "dodge") +
  theme_scoringutils() + 
  coord_cartesian(ylim = c(0, 1)) +
  labs(y = "Corr. ranks single forecasts", x = "Forecast horizon")

# p_cor_scores <- cor_scores |> 
#   mutate(`Target type` = target_type) |>
#   ggplot(aes(x = horizon, y = cor, 
#              fill = `Target type`, color = `Target type`)) + 
#   geom_histogram(stat = "identity", position = "dodge") +
#   theme_scoringutils() + 
#   coord_cartesian(ylim = c(0, 1)) +
#   labs(y = "Spearman corr. scores single forecasts", x = "Forecast horizon")

p_cor_skill <- correlation_rel_skill |>
  rename(`Target type` = target_type) |>
  ggplot(aes(x = horizon, y = cor, fill = `Target type`, color = `Target type` )) + 
  geom_histogram(stat = "identity", position = "dodge") +
  theme_scoringutils() + 
  coord_cartesian(ylim = c(0, 1)) +
  labs(y = "Correlation rel. skill", x = "Forecast horizon")

plot_cors <- p_cor_ranks  + p_cor_skill +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom")

ggsave(filename = "output/figures/HUB-cors.png", width = 7, height = 3, 
       plot = plot_cors)

@@

```{r}

# wis_vs_total(scores) + 
#   labs(y = "Mean WIS")
# ggsave("output/figures/HUB-mean-scores-vs-total.png", width = 7, height = 5)
# 
# wis_vs_total(scores, summary_fct = stats::sd) + 
#   labs(y = "Sd WIS")
# ggsave("output/figures/HUB-sd-scores-vs-total.png", width = 7, height = 5)

# wis_vs_total(scores, summary_fct = stats::sd) + 
#   labs(y = "Sd WIS") + 
#   scale_x_continuous(labels = scale_fn, trans = "log10") +
#   scale_y_continuous(labels = scale_fn, trans = "log10") 
# ggsave("output/figures/HUB-sd-scores-vs-total-log-log.png", width = 7, height = 5)

### 2022-08-02 main.tex
update manuscript

As an empirical example for evaluating forecasts on the natural and on the log scale we use forecasts from the European Forecast Hub \citep{europeancovid-19forecasthubEuropeanCovid19Forecast2021, sherrattPredictivePerformanceMultimodel2022}. Every week the European Forecast Hub collates and aggregates forecasts for different COVID-19 related targets from teams around the world. Forecasts are made one to four weeks ahead into the future and follow a quantile-based format with a set of 22 quantiles plus the median ($0.01, 0.025, 0.05, ..., 0.5, ... 0.95, 0.975, 0.99$). The forecasts for the purpose of this illustrations are two-week-ahead forecasts made between XXX and XXX for reported cases and deaths from COVID-19 in XXX countries. Where not otherwise stated, we report results for a two-week-ahead forecast horizon.

@@

Variation of scores across targets (see Figure XX) and locations (see Figure \ref{fig:HUB-mean-locations}A, B) is much larger on the natural than on the log scale, where scores are more evenly distributed. On the natural scale, average scores per location correlate strongly (XX) with the average number of observed cases or deaths (See Figure XX). On the log scale, this correlation is much weaker and slightly negative. Forecasts across different locations receive more similar scores, with a tendency for scores to be slightly higher in locations with few observed cases and deaths (see Figures XX and \ref{fig:HUB-mean-locations}. 
%This is in line with the observation that the association between variance is smaller than mu^2 + mu, meaning that we would expect this. 
For individual forecasts, rankings between models are mostly preserved (correlation: X for cases, Y for deaths), with differences increasing across forecast horizons (see Figure \ref{fig:HUB-cors-scores}).

The correlation between scores on the log and natural scale for different locations is XX for cases and XX for deaths. 
The correlation between the actual observations is 0.628 for cases and 0.631 for deaths. IS THAT HELPFUL?

\begin{table}[h!]]
    \centering

    \begin{tabular}{lllcc}
    \toprule
    target\_type & quantity & measure & natural & log\\
    \midrule
    Cases & Observations & mean & 20308 & 8.39\\
    Cases & Observations & sd & 43405 & 2.11\\
    Cases & Observations & var & 1884002308 & 4.46\\
    Deaths & Observations & mean & 259 & 3.79\\
    Deaths & Observations & sd & 532 & 2.11\\
    \addlinespace
    Deaths & Observations & var & 283492 & 4.46\\
    Cases & WIS & mean & 4544 & 0.28\\
    Cases & WIS & sd & 17585 & 0.49\\
    Deaths & WIS & mean & 28 & 0.22\\
    Deaths & WIS & sd & 67 & 0.26\\
    \bottomrule
    \end{tabular}

    \caption{CAPTION}
    \label{tab:HUB-summary}
\end{table}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\textwidth]{output/figures/HUB-cors-skill-rank.png}
    \caption{Correlation between scores on the natural and on the log scale. A: Correlation between relative ranks (the rank in terms of WIS divided by the number of models) for individual forecasts, separated by forecast horizon. B: Correlation between relative skill values achieved by models across locations and time points, separated by forecast horizon.}
    \label{fig:HUB-cors-scores}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\textwidth]{output/figures/HUB-ranking-diffs.png}
    \caption{Distribution of rankings of the different Hub models. Rankings are computed as relative rankings, i.e. the ranking divided by the total number of models available for a given forecast target.}
    \label{fig:HUB-rankings}
\end{figure}

@@

Both on the natural scale as well as on the log scale, scores increase considerably with increasing forecast horizon (see Figure \ref{fig:HUB-scores-horizon}). The increase is more pronounced for cases than for deaths and arguably represents an increase in the relative difficulty to forecast values further into the future. A similar increase both on the natural and the log scale indicates that we are able to observe the increase in difficulty regardless of how we evaluate the scores.

>>>>

As an empirical example for evaluating forecasts on the natural and on the log scale we use forecasts from the European Forecast Hub \citep{europeancovid-19forecasthubEuropeanCovid19Forecast2021, sherrattPredictivePerformanceMultimodel2022}. Every week the European Forecast Hub collates and aggregates forecasts for different COVID-19 related targets from teams around the world. Forecasts are made one to four weeks ahead into the future and follow a quantile-based format with a set of 22 quantiles plus the median ($0.01, 0.025, 0.05, ..., 0.5, ... 0.95, 0.975, 0.99$). The forecasts for the purpose of this illustrations are two-week-ahead forecasts made between XXX and XXX for reported cases and deaths from COVID-19 in XXX countries. We filtered all forecasts submitted to the Hub to only include models which have submitted forecasts for both deaths and cases for 4 horizons in 32 locations on at least 16 forecast dates. Where not otherwise stated, we report results for a two-week-ahead forecast horizon.

\begin{table}[h!]
    \centering

    \begin{tabular}{lllcc}
    \toprule
    target\_type & quantity & measure & natural & log\\
    \midrule
    Cases & Observations & mean & 20308 & 8.39\\
    Cases & Observations & sd & 43405 & 2.11\\
    Cases & Observations & var & 1884002308 & 4.46\\
    \addlinespace
    Deaths & Observations & mean & 259 & 3.79\\
    Deaths & Observations & sd & 532 & 2.11\\
    Deaths & Observations & var & 283492 & 4.46\\
    \addlinespace
    \hline
    \addlinespace
    Cases & WIS & mean & 4544 & 0.28\\
    Cases & WIS & sd & 17585 & 0.49\\
    \addlinespace
    Deaths & WIS & mean & 28 & 0.22\\
    Deaths & WIS & sd & 67 & 0.26\\
    \bottomrule
    \end{tabular}

    \caption{Table with summary statistics for observations and scores for forecasts from the ECDC data set.}
    \label{tab:HUB-summary}
\end{table}

@@

Variation of scores across targets (see Figure \ref{fig:HUBXXX} and locations (see Figure \ref{fig:HUB-mean-locations}A, B) is much larger on the natural than on the log scale, where scores are more evenly distributed. On the natural scale, average scores per location correlate strongly (XX) with the average number of observed cases or deaths (See Figure XX). On the log scale, this correlation is much weaker and slightly negative. Forecasts across different locations receive more similar scores, with a tendency for scores to be slightly higher in locations with few observed cases and deaths (see Figures XX and \ref{fig:HUB-mean-locations}. 
%This is in line with the observation that the association between variance is smaller than mu^2 + mu, meaning that we would expect this. 
For individual forecasts, rankings between models are mostly preserved (correlation: X for cases, Y for deaths), with differences increasing across forecast horizons (see Figure \ref{fig:HUB-cors}).

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\textwidth]{output/figures/HUB-cors.png}
    \caption{Correlation between scores on the natural and on the log scale. A: Correlation between relative ranks (the rank in terms of WIS divided by the number of models) for individual forecasts, separated by forecast horizon. B: Correlation between relative skill values achieved by models across locations and time points, separated by forecast horizon. SHOULD THIS BE SPEARMAN?}
    \label{fig:HUB-cors}
\end{figure}

@@

Both on the natural scale as well as on the log scale, scores increase considerably with increasing forecast horizon (see Figure \ref{fig:HUB-scores-horizon}). The increase is more pronounced for cases than for deaths and arguably represents an increase in the relative difficulty to forecast values further into the future. A similar increase both on the natural and the log scale suggests that we are able to observe the increase in difficulty regardless of how we evaluate the scores.
